---
title: "Reproducing Maine Ranked Choice Tabulations"
author: "Ken Lane"
date: "12/23/2019"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    code_folding: show
    theme: united
    highlight: tango
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
urlcolor: red
includes:
  in_header: preamble.tex
bibliography: rcvr.bib
link-citations: yes
editor_options: 
  chunk_output_type: inline
---

# Introduction
## Ranked choice voting in Maine
Commencing in June 2018,
the State of Maine has used a type of ranked choice voting (RC) commonly called instant runoff (RC/IR).
In November 2018
Maine became
the first state to implement
RC/IR in elections
for the United States House of Representatives
and United States Senate.
The use of RC was decisive in one congressional contest, 
with CD2 incumbent Bruce Poliquin
receiving the plurality (43.66\%) of first round votes,
but lacking a majority, lost in the second round of tabulation [@nyt-poliquin].

Subsequent to the November election, Congressman Poliquin and others filed a lawsuit against Maine Secretary of State Matthew Dunlap, 
seeking to be declared the CD2 winner by virtue of a plurality in the first round tabulations [@npr-poliquin].
Congressman Poliquin and fellow plaintiffs maintained that

>
    "... the RCV Act is unconstitutional, both facially and as applied, and that it violates the Voting
    Rights Act. They maintain that the ballot form and instructions were too confusing and that the
    manner by which Defendant Dunlap tabulated the votes diluted the votes cast by Poliquin supporters
    and otherwise disenfranchised too many Maine voters to withstand scrutiny."


While unsuccessful, the Poliquin lawsuit brought attention to the tabulation of
RC/IR ballots.
Public criticism included references to a ``proprietary
algorithm'' used by the State to mechanize runoff tabulations [@muszynski].
The algorithm in question is implemented in proprietary software,
and as such,
is not available for public scrutiny.
Nonetheless,
shortly after the public release of cast vote records on November 21, 2018,
multiple parties reported successful independent duplication of official State
tabulations [@mpr-poliquin],
pointing out

>
    "... that the software the state uses is proprietary.
    But the rules that are used by that software are just as public
    as the election results."


It would seem that independent duplication of official
results falls short of validating
the **propriety** of proprietary software.
In particular, the following questions remain:

1. Are the **rules complete**?
       Do the rules completely
       determine the disposition of every ballot?
       If not, proprietary software would seem inappropriate.
       
2. Are the **rules consistent**? If not, every 
      implementation will fall short and a revision of the
      rules is indicated.

3. Are the **rules independent**?
      Would a proper subset of the rules always produce
      the same result?
      Independence is a component of clarity and concision,
      key attributes for the communication and implementation
      of the rules.

4. Does the **implementation conform** to the rules?
      Independent implementations often share share common flaws.
      Duplication of results for a single election is inconclusive.

The validation of software implementations is an extraordinary complex
enterprise and has been carefully considered elsewhere [@vvsg; @vvsgd2; @eac-certification].
The focus here is the rules.

## Approach
The case study follows from the author's effort to independently and unofficially
validate RC/IR tabulations presented by State of Maine for the three RC/IR contests.
In particular, the goal was to do the following:

1. Create a mathematical model of tabulation processing
     rendered in the vocabulary of set theory.
     
2. Examine State of Maine RC/IR election rules in the context of the model.

3. Develop and implement an algorithm to independently verify (potentially) tabulations
produced by the proprietary algorithm currently deployed in State of Maine RC/IR elections. 

To accomplish the first two tasks,
only official and publicly available sources were used to identify RC/IR tabulation rules.
To avoid adopting hidden assumptions,
others with potential knowledge of rules or tabulation implementations were not consulted.
A "clean room" strategy likewise affords
an opportunity to evaluate the independence (clarity
and concision) embodied by the rules.
The assumption being:

>
   Official rules ought to be sufficient to guide a software implementation
        of an **independent tabulator** capable of reproducing Maine RC/IR tabulations.

The results of the modeling effort are provided in a seperate
[document](preference_extraction.pdf).
An R package `rcvr` (Recievor) has been written to accomplish the final task
is demonstrated in the sections that follow.

# Recievor
Minimal packages required for `rcvr` are `plyr` and `tidyverse`
[^1].

[^1]: `kableExtra` loaded soley to render this document and is not
needed by Recievor.


```{r setup, include = TRUE, echo = TRUE, message = FALSE}
  library(plyr)
  library(tidyverse)
  library(kableExtra)
  library(rcvr)
```

## Contest database
Metadata to reconstruct elections are stored in the textfile `contest_keys.txt`,
including filenames and URLs for downloading original
cast vote record files from
the Maine Office of the Secretary of State, Division
of Elections [@ME_results].
Thus far, three data sets are available, including:

- June 12, 2018 - Primary Election
  - Governor - Democrat (132250 CVRs)
  - Representative to Congress District 2 - Democrat (50845 CVRs)
  
- November 6, 2018 - General Election
  - Representative to Congress District 2 (296077 CVRs)


```{r}
  dat <- read.table("contest_keys.txt",
                    sep = "|",
                    col.names = c(
                      "object",
                      "field",
                      "value"
                    ),
                    stringsAsFactors = FALSE,
                    comment.char = "#")
  knitr::kable(dat,
               row.names = FALSE,
               caption   = "Contest Keys Database") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                  full_width = FALSE) %>%
    scroll_box(width = "100%", height = "300px")
```

## Contest instances
Recievor models each election using an S3 class `contest`
[^2].

[^2]: See `contest()` manual page [here](html/contest.html).

Use `rrcv::contest()` to construct a `contest` instance,
which will store metadata as `attributes`
and edited ballot data in an underlying
dataframe[^3].

[^3]: See `clean_cvr_alternatives()` manual page [here](html/clean_cvr_alternatives.html)
and `standardize_rankings` [here](html/standardize_rankings.html).

1. If `original == TRUE` the `contest_str` must be set to an `object`
found in the contest database,
as well as valid values for `source`, `CVR_file`,
and `alternative` fields
matching `object`.

2. If `original == FALSE` parameters for a folder (`source =`)
and a list of cast vote records (`CVR_files =`)
must be supplied when invoking `rcvr::contest()`.

Once a `contest` instance is created tabulation is
accomplished using `rcvr::conduct_runoff()`.

## Tabulation options
Tabulation of `contest` instances is accomplished by `rcvr::conduct_runoff()`,
which provides options for processing
overvotes and undervotes [^4].

[^4]: See manual page for `conduct_runoff()` [here](html/conduct_runoff.html).

For example under Maine rules,
an overvote at a particular rank
invalidates that choice and all subsequent choices.
Other jurisdictions may ignore
any or all overvotes.
Under Maine rules, single
undervotes are ignored but **two consecutive**
undervotes invalidate all subsequent choice [@250c535].
Maine law is more restrictive,
with **two sequential** undervotes
invalidating subsequent choices[@21AMRS723A].
The default for `rcvr::conduct_runoff()`

- `sequentialU == TRUE`
- `max_U = 2`
- `sequentialO == TRUE`
- `max_O  = 1`

# Retabulations
## June 2018 Democratic Primary for Governor
Call the constructor and encode the ballot choices to append
to random ballot displayed in table below.

$~$

```{r cache = TRUE}
  gov_June <- rcvr::contest(
                      contest_name = "Democratic Primary for Governor",
                      original     = TRUE,
                      contest_str  = "gov_June",
                      creator      = "Ken Lane")
  codes    <- rcvr::get_codes(gov_June)
```

```{r}  
  kable(rcvr::show_contest(gov_June),
          row.names = FALSE,
          booktabs  = TRUE,
          caption   = "Metadata: Democratic Primary for Governor",
          col.names = c("Item", "Value"),
          digits    = 2,
          longtable = TRUE) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                  full_width        = FALSE) %>%
    scroll_box(width = "100%", height = "300px")
```

```{r}
  num_ballots <- attr(gov_June, "num_records")
  df <- as.data.frame(gov_June)
  df <- mutate(df, code = codes)
  knitr::kable(df[sample(1:num_ballots, 100, replace = FALSE),],
               row.names = FALSE,
               caption   = "100 Random Ballots: Democratic Primary for Governor") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "300px")
```

$~$

June 2018 gubernatorial primary results posted by the State of Maine
can be found [here](https://www.maine.gov/sos/cec/elec/results/results18.html#june18).
This tabulation cannot be reproduced using the process described
by the administrative rules
[@250c535].
The tabulation corresponding to the rules is easily generated.

```{r cache = TRUE}
tab1 <- rcvr::conduct_runoff(gov_June)
df1  <- rcvr::make_report4(tab1)
```


Official Maine results can be duplicated by "tweaking" the rules [^5].

[^5]: Tweaking implemented in `tabulate_cvr_rank()`.
See manual page [here](html/tabulate_cvr_rank.html).

```{r cache = TRUE}
tab2 <- rcvr::conduct_runoff(gov_June,
                             sequentialU = TRUE,
                             june2018    = TRUE)
df2  <- rcvr::make_report4(tab2)
```


```{r echo = FALSE}
  df  <- cbind(df1[,1:2], df2[,2], df1[,3], df2[,3], df1[,4], df2[,4], df1[,5], df2[,5])
  knitr::kable(df, 
             row.names = FALSE,
             booktabs  = TRUE,
             caption   = "June 2018 Democratic Primary for Governor",
             col.names = c("", "Rules", "Reported", "Rules", "Reported", "Rules", "Reported", "Rules", "Reported"),
             longtable = TRUE
) %>%
  kable_styling(bootstrap_options = c("striped", "Scale_down", "repeat_header"),
                full_width        = FALSE) %>%
  add_header_above(c(" ", "Round 1" = 2, "Round 2" = 2, "Round 3" = 2, "Round 4" = 2)) %>%
  pack_rows("Candidate Names", 1, 8, latex_gap_space = "2em") %>%
  pack_rows("Ballot Exhausted", 9, 11, latex_gap_space = "2em")
```

While internally consistent (non-contradictory),
the administrative rules are not
in alignment with the corresponding statute
with regard to skipped rankings [@21AMRS723A].
In fact, parties reporting duplicating
November CD2 tabulations were unable to duplicate
results for the June Gubernatorial primary use the
same algorithm [^6].

[^6]: Personal communication with Nathan Tefft. See his
python implementation [here](https://github.com/ntefft/rcv).

Three idiosyncratic ballots from 
the June 2018 gubernatorial primary
have been identified.

```{r ehco = TRUE}
r1   <- unlist(tab1$round4$result)
r2   <- unlist(tab2$round4$result)
diff <- (r1 != r2)
df   <- gov_June %>% as.data.frame()
df   <- mutate(df, code = codes)
dat  <- df[diff, c(1, 2, 12)]
knitr::kable(dat, 
             row.names = FALSE,
             booktabs  = TRUE,
             caption   = "Ballots of Interest",
             col.names = c("CVR ID", "Precinct", "Coded Choices"),
             digits    = 2,
             longtable = TRUE)  %>%
  kable_styling(bootstrap_options = c("striped", "repeat_header"),
                full_width        = FALSE)
```

The three ballots would be awarded to candidate E (Janet Mills) in
the fourth round using the "two consecutive" trimming rule.
Independent implementations report 3 fewer votes for Janet Mills
in the final round than the official tabulation.
The results generated by the proprietary algorithm can be
achieved by trimming each ballot after the second skipped ranking,
**except** where the first skipped ranking is
the first choice ranking.
There were three such exceptional ballots

- File: govd-2.xlsx, Row: 1315, Code: UGUBAEDC, Cast Vote Record: 13216: Precinct: Old Orchard Beach
- File: govd-2.xlsx, Row: 33534, Code: UGUECGAF, Cast Vote Record: 141592: Precinct: Eastport
- File: govd-2.xlsx, Row: 41151, Code: UCUAUUUU, Cast Vote Record: 173959: Precinct: Harpswell W1

## June 2018 Maine CD2 Democratic Primary
Use `contest.R` to construct a `contest` instance.

```{r cache = TRUE}
  d2_June <- rcvr::contest(
    contest_name = "2018 Maine CD2 Democratic Primary",
    original     = TRUE,
    contest_str  = "d2_June",
    creator      = "Ken Lane",
    quietly      = TRUE)
  codes    <- rcvr::get_codes(d2_June)
```

```{r echo = FALSE}  
  knitr::kable(rcvr::show_contest(d2_June), 
                row.names = FALSE,
                booktabs  = TRUE,
                caption   = "Metadata: CD2 Democratic Primary",
                col.names = c("Item", "Value"),
                digits    = 2,
                longtable = TRUE)  %>%
              kable_styling(bootstrap_options = c("striped", "repeat_header"),
                            full_width        = FALSE) %>%
    scroll_box(width = "100%", height = "300px")
```

```{r}
  num_ballots <- attr(d2_June, "num_records")
  df <- as.data.frame(d2_June)
  df <- mutate(df, code = codes)
  knitr::kable(df[sample(1:num_ballots, 100, replace = FALSE),],
               row.names = FALSE,
               caption   = "100 Random Ballots: CD2 Democratic Primary") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "300px")
```

$~$

Second district results can be reproduced using the administrative rules.
Fewer ranks (5) avoid the rare pathologies found in
the gubernatorial primary. In fact, the "tweaked" algorithm
produces the same result as the official tabulation
and the rules.
First, using the administrative rules.

```{r cache = TRUE}
tab1 <- rcvr::conduct_runoff(d2_June)
df3 <- make_report2(tab1)
```

And now invoke the "tweaked" version.

```{r cache = TRUE}
tab2 <- rcvr::conduct_runoff(
  d2_June,
  sequentialU = TRUE,
  june2018    = TRUE)
df4 <- make_report2(tab2)
df  <- cbind(df3[,1:2], df4[,2], df3[,3], df4[,3])
```

```{r echo = FALSE}
knitr::kable(df, 
             row.names = FALSE,
             booktabs  = TRUE,
             digits    = 2,
             caption   = "June 2018 Maine CD2 Democratic Primary",
             col.names = c("", "Rules", "Reported", "Rules", "Reported"),
             longtable = TRUE
) %>%
  kable_styling(bootstrap_options = c("striped", "repeat_header"),
                full_width        = FALSE) %>%
  add_header_above(c(" " = 1, "Round 1" = 2, "Round 2" = 2)) %>%
  pack_rows("Candidate Names", 1, 4, latex_gap_space = "2em") %>%
  pack_rows("Ballot Exhausted", 5, 7, latex_gap_space = "2em")

```

## November 2018 CD2 General Election
Use `contest.R` to construct a `contest` instance.

```{r cache = TRUE}
  d2_Nov <- rcvr::contest(
                      contest_name = "November 2018 Maine CD2 General Election",
                      original     = TRUE,
                      contest_str  = "d2_Nov",
                      creator      = "Ken Lane",
                      quietly      = TRUE)
  codes <- get_codes(d2_Nov)
```

```{r echo = FALSE}  
  knitr::kable(rcvr::show_contest(d2_Nov), 
                row.names = FALSE,
                booktabs  = TRUE,
                caption   = "Metadata: CD2 General Election",
                col.names = c("Item", "Value"),
                digits    = 2,
                longtable = TRUE)  %>%
              kable_styling(bootstrap_options = c("striped"),
                            full_width        = FALSE) %>%
    scroll_box(width = "100%", height = "300px")
```

```{r}
  num_ballots <- attr(d2_Nov, "num_records")
  df <- as.data.frame(d2_Nov)
  df <- mutate(df, code = codes)
  knitr::kable(df[sample(1:num_ballots, 100, replace = FALSE),],
               row.names = FALSE,
               caption   = "100 Random Ballots: CD2 General Election") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "300px")
```

$~$

Tabulations using the administrative rules agree with official results [see here](https://www.maine.gov/sos/cec/elec/results/results18.html#Nov6).

```{r cache = TRUE}
tab <- rcvr::conduct_runoff(d2_Nov)
df  <- make_report2(tab)
```

```{r echo = FALSE}
knitr::kable(df, 
             row.names = FALSE,
             booktabs  = TRUE,
             digits    = 2,
             caption   = "November 2018 Maine CD2 General Election",
             col.names = c(" ", "Round 1", "Round 2")) %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width        = FALSE) %>%
  pack_rows("Candidate Names", 1, 4, latex_gap_space = "2em") %>%
  pack_rows("Ballot Exhausted", 5, 7, latex_gap_space = "2em")

```

# Session Information
```{r echo = TRUE}
  xfun::session_info(dependencies = FALSE)
```

# Bibliography